Simulations of gravity-driven flow of binary liquids in microchannels

### 2011-01-06 Text/paper.tex
Added Mostowfi revisions. Jos revisions are not yet finished. Need to polish figures.

\title{The binary liquid model for gravity-driven microchannel simulations}

@@

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[uofa]{J.~Derksen}
\ead{kuzmin@ualberta.ca}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

\begin{abstract}
This work shows that the free-energy binary liquid lattice Boltzmann method is able to
reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
capillary
numbers
$\Ca$ (defined in Eq.~\ref{capillary:number:definition}), the flow is thoroughly described by
Bretherton \cite{bretherton},
with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
from the Bretherton correlation for $\Ca\geq 0.005$, studied experimentally and numerically. This
work aims to examine the applicability of the lattice Boltzmann binary liquid model to simulations
of bubble flow in liquids for a range of $\Ca$ numbers. The binary
liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
It was found that the interface thickness needs to be $50$
percent of the film thickness for the simulation to give convergent results. When
the convergence criterion is fulfilled, the results are in a good agreement with the
numerical simulations of \citet{giavedoni-numerical} in the range of $\Ca$ from $0.01$ to $1$. 
\end{abstract}

@@

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow describes the flow of long bubbles in
microchannels. It was found that the front meniscus film thickness is proportional
to $\Ca^{2/3}$ for capillary numbers smaller than $0.003$, where the capillary number is defined as:
\begin{equation}
\label{capillary:number:definition}
\Ca=\frac{\mu_\mathrm{liq} U_\mathrm{bubble}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $U_{bubble}$ is the bubble velocity, and
$\gamma$ is the surface tension between gas and liquid.

While there are a number
of works which simulate the Bretherton problem by the boundary value approach
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion, or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is spread over several grid nodes, the question of
proper film resolution in comparison with the interface resolution arises.  This work is focused on
resolving the parameter range of the binary liquid lattice
Boltzmann method to validate it for a range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries and allows incorporation of
physical phenomena on the microscopic level, as in the case of multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface using continuous interface methods where the interface spans over several grid nodes.
Such a representation
brings issues of the film thickness resolution versus interface
resolution -- the diffuse interface should be resolved in such a way as to have
negligible effect on the physics of the film.

The binary liquid free-energy LB model \cite{swift} we used
simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
neglected. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effect on the film thickness for a relatively wide range of Reynolds
numbers.
Therefore, the major governing factor for microchannel flows is not the density ratio, but the
viscosity ratio. Moreover, the results based on the uniform density, as in the case of the LBM
binary liquid model, are in good agreement with other simulations
\cite{giavedoni-numerical,heil-bretherton} in which inertia effects were taken into account.

The goal of this paper is to understand how to choose the
binary liquid model parameters, the viscosity ratio and the grid
resolution to obtain correct film thicknesses.

The paper is organized as follows.  First, we briefly
explain the binary liquid lattice Boltzmann model and review the literature
for the Bretherton problem. Then, the parameters involved in the
simulation are examined and presented in the results section. The paper is
concluded with a summary of the main findings.

\section{Bretherton bubble flow}
Bretherton \cite{bretherton} studied a long bubble
moving in a tube filled with liquid. It was found that the deposition film thickness
is proportional to $\Ca^{2/3}$ in the range of small capillary numbers. Later on
it was realized \cite{wong-films,wong-pressure} that the film thickness
is proportional to $\Ca^{2/3}$ only in the certain region behind the front meniscus and
that the film thickness
varies over the bubble length for bubbles of finite length. Numerical
simulations \cite{giavedoni-numerical} and experimental studies
\cite{kreutzer-pressure-drop} showed a deviation from the $\Ca^{2/3}$ rule and Reynolds number 
effect on the film thickness for capillary numbers larger than $0.003$.
To  predict consistently a flow pattern for capillaries in
different ranges of parameters, simulations have been conducted and have been typically validated with the
small capillary numbers Bretherton problem.

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} also used the finite volume method to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. \citeauthor{wong-films} \cite{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and showed calculated different shapes in the
slug cross sections and menisci appearance.
\citet{heil-bretherton,ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using the finite element method. \citet{giavedoni-numerical} performed cross validation of the
finite element solution with previously published results.
The solutions were obtained for circular and planar cases.

@@

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for a relatively large capillary numbers, and
the authors found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}.

Resolving the interface on a fine level for capillary numbers smaller than
$0.003$ is computationally expensive even in the two-dimensional case
(the amount of memory necessary to perform the simulation is of the order
of tens or hundreds gigabytes).  This work is therefore based on a
comparison of the results obtained with the known data
validated for the Bretherton bubble flow.  In this paper, we present
techniques for initialization of the simulations and discuss optimal
parameter ranges for the binary liquid LB model.

@@

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), i=1\div8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \frac{\phi c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi^m
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), i=1\div8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model.  The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$ with
the sound speed parameter $c_s^2=1/3$.  The weights related to the
inclusion of the surface tension coefficient into the equations are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}=
-\partial_{\alpha}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity and
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter.

The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken as linearly dependent on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}\tau_{\mathrm{liq}}$.
The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k A}{9}}$.

\section{Lattice Boltzmann benchmark}
To properly simulate and compare simulation results with the data
published in the literature, one needs to design a numerical benchmark addressing
the following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffuse nature of the interface in the multiphase model should not
influence the film thickness results. This is obviously achieved only for certain
interface resolutions relative to the film thickness. The same applies to spurious
currents arising due to surface tension discretization, which
can influence mass transfer in the thin film, thereby compromising the
overall solution. Therefore, the diffuse interface should not noticeably affect the film
thickness.
 \item The wettability coefficient which in principle can control the
dynamic angle and the menisci appearance \cite{pagonabarraga-finger} should not
influence the overall solution.
 \item The free-energy model examined in this work is a single density, two
viscosities model. The classical Bretherton problem is a free-interface problem between liquid and
gas
where the film thickness is established by itself. Thus, the viscosity and the density ratios for
the
classical
Bretherton problem should
be large. However, in terms of the lattice Boltzmann binary liquid framework only the viscosity
ratio can be addressed. Our computations show that the results are consistent with other works and
inertial effects can be neglected.
  \item The Bretherton problem addresses a bubble flow driven by a pressure
difference, not a body force. However, to the best authors' knowledge the pressure boundary
conditions are still not yet developed for the binary liquid model, because it involves the
treatment of associated gradients and Laplacians on the boundary.
This kind of pressure boundary conditions for the binary liquid model is currently
under development, and preliminary results show that the differences between flows driven by a body
force and by
pressure difference are small as far as the film
 thickness is of interest.  In this work we limit ourselves to the study
of body force driven flows, because of their simplicity and better numerical stability. This
case implies the periodic boundary conditions. As soon as the periodic boundary
conditions are applied, not a single bubble but whole bubble train is simulated. In this case one
needs
to ensure that the distance between bubbles is large enough to exclude mutual bubble influence.
 \item Most simulations are done for a channel with circular cross section,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework, where one needs to calculate the Laplacian and gradients near boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed specifically for curved boundaries,
the analysis of error introduced by the boundary conditions becomes an issue. Though the circular
axis-symmetric case can be simulated on a rectangular grid by introducing specific mass and force
terms in the continuity and Navier-Stokes equation \cite{halliday-circular}, the lattice Boltzmann
incorporation of force and mass terms is in the process of development and validation. Therefore,
the benchmark is chosen as a plane-symmetric two-dimensional case.

\end{itemize}
Given all the concerns and challenges, the suggested lattice Boltzmann framework
is a two-dimensional flow driven by a body force.  An analytical
solution to this problem is not known for a wide range of capillary
numbers.
Though some authors \cite{sehgal-microchannel} use the gravity driven model to
validate the Bretherton film thickness, the latter should be used with caution
since the velocity of the bubble is different from the fluid velocity and the
bubble shape
is different in the front and rear menisci depending on Bond number ($\mathrm{Bo}=\frac{\Delta \rho g
l^2}{\gamma}$). \citet{wong-films}
analytically described the film thickness variation along a bubble varying from $\Ca^{1/2}$ to
$\Ca^{2/3}$. The bubble should be long enough to have the film thickness proportional to
$\Ca^{2/3}$. Therefore, we chose the bubble length to be $5$ channel heights. This number is shown
to be sufficient to give results consistent with the theory. The
channel length is taken to be $3$ bubble lengths (or $15$ channel heights)
to minimize influence of one bubble on another,
because of the periodicity of boundary conditions.
Figure \ref{fig:benchmark:sketch} is a sketch of the geometry
used for the benchmark.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/benchmark.eps}
\caption{The benchmark sketch. $\delta$ corresponds to film thickness. $\xi$ corresponds to the interface thickness. We take the microchannel length to be 15 times larger than its height.
\label{fig:benchmark:sketch}}
\end{figure}

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
We now introduce the procedure of nondimensionalization, as well as
the initialization technique we used to determine the parameters necessary for
the simulation. The capillary number defined in Eq.~\ref{capillary:number:definition} governs the interface thickness and is.
The bubble velocity $U_{\mathrm{bubble}}$ is taken in the center of the bubble where the film
thickness is examined.
While parameters such as $\mu_{\mathrm{liq}}$, $\gamma$ and the capillary number $\Ca$ are given
or defined through the model parameters, the bubble velocity $U_{\mathrm{bubble}}$ is not given
explicitly.
In our simulations we approximately match and initialize the velocity with the
Poiseuille profile center velocity, which is given as:
\begin{equation}
\begin{aligned}
U_{\mathrm{bubble}} \approx U_{\mathrm{liq}}=\frac{H_{\mathrm{eff}}^2}{8
\mu_{\mathrm{liq}}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{\mathrm{bubble}}\approx
\frac{{(N_y-2)}^2}{\frac{8}{6}(\tau_{\mathrm{liq}}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
where $N_y$ is the grid number in the vertical direction. As will be discussed further, the
effective channel width is defined as $H_{\mathrm{eff}}=N_y-2$ because of bounce-back nodes.
For the Poiseuille profile one can estimate the body force (pressure gradient) through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{(N_y-2)}^2}\gamma \Ca.
\end{aligned}
\end{equation}
The body force (\ref{poiseuille:velocity:center}) causes the bubble
velocity to differ from that implied by the capillary number because of the bubble presence in a flow. That is
why the iteration proceduer is needed to achieve the 
desired velocity and 
capillary number. 
However, the initialization body force given by (\ref{poiseuille:velocity:center}) is a good
starting approximation.

>>>>

\title{Simulations of gravity-driven flow of binary liquids in microchannels}

@@

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

\begin{abstract}
%This work shows that the free-energy binary liquid lattice Boltzmann method is able to
%reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
%capillary
%numbers
%$\Ca$ (defined in Eq.~\ref{capillary:number:definition}), the flow is thoroughly described by
%Bretherton \cite{bretherton},
%with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
%from the Bretherton correlation for $\Ca\geq 0.005$, studied experimentally and numerically. This
%work aims to examine the applicability of the lattice Boltzmann binary liquid model to simulations
%of bubble flow in liquids for a range of $\Ca$ numbers. The binary
%liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
%It was found that the interface thickness needs to be $50$
%percent of the film thickness for the simulation to give convergent results. When
%the convergence criterion is fulfilled, the results are in a good agreement with the
%numerical simulations of \citet{giavedoni-numerical} in the range of $\Ca$ from $0.01$ to $1$. 
In this work a free-energy binary liquid lattice-Boltzmann scheme is used to simulate
Taylor/Bretherton flow in a micro-channel where elongated gas bubbles move through a liquid with
thin liquid films between the bubbles and the channel walls. The numerical scheme has a diffuse
interface, and a main focus of our work is to assess resolution requirements for correctly resolving
the liquid film and bubble motion. The simulations are two-dimensional and span a capillary number
range of $0.03$ to $0.8$ where the capillary number is based on the liquid’s dynamic viscosity, the
velocity of the bubble, and the interfacial tension. The flow is driven by a body force, and
periodic boundary conditions apply in the streamwise direction. We obtain grid independent results
as long as the liquid film thickness is at least two times the width of the diffuse interface, with
film thicknesses in accordance to literature results. We also show that the results in terms of film
thicknesses are largely insensitive to the liquid-gas viscosity ratio and wettability parameters. 
\end{abstract}

@@

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow deals with long gas bubbles moving through liquid in
narrow channels. It was found that the front meniscus film thickness is proportional
to $\Ca^{2/3}$ for capillary numbers smaller than $0.003$, where the capillary number is defined as:
\begin{equation}
\label{capillary:number:definition}
\Ca=\frac{\mu_\mathrm{liq} U_\mathrm{bubble}}{\gamma},
\end{equation}
where $\mu_{liq}$ is the liquid viscosity, $U_{bubble}$ is the bubble velocity, and
$\gamma$ is the interfacial tension between gas and liquid.

While there are many works which simulate the Bretherton problem by the boundary value approach
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion, or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is smeared our over several grid nodes, the question of
proper film resolution in comparison with the interface resolution arises.  This work is focused on
exploring the parameter range of the binary liquid lattice
Boltzmann method for correctly resolving the flow in a range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries and allows incorporation of
physical phenomena on the microscopic level, as in the case of multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface using continuous interface methods where the interface spans over several grid nodes.
Such a representation
brings issues of the film thickness resolution versus interface
resolution -- the diffuse interface should be dealt with in such a way as to have a
negligible effect on the physics of the film.

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
neglected. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers.
Therefore, the major governing factor for microchannel flows is not the density ratio, but the
viscosity ratio. Moreover, the results based on the uniform density, as in the case of the LBM
binary liquid model, are in good agreement with other simulations
\cite{giavedoni-numerical,heil-bretherton} in which inertia effects were taken into account.

The goal of this paper is to understand how to choose the
binary liquid model parameters, the viscosity ratio and the grid
resolution to obtain correct flow physics with an emphasis on film thicknesses.

The paper is organized as follows.  First, we review the literature
for the Bretherton problem and briefly
explain the binary liquid lattice Boltzmann model. Then, the parameters involved in the
simulation are examined and presented in the results section. The paper is
concluded with a summary of the main findings.

\section{Bretherton bubble flow}
Bretherton \cite{bretherton} studied a long bubble
moving in a tube filled with liquid. It was found that the deposition film thickness
is proportional to $\Ca^{2/3}$ in the range of small capillary numbers. Later,
it was realized \cite{wong-films,wong-pressure} that the film thickness
is proportional to $\Ca^{2/3}$ only in the certain region behind the front meniscus and
that the film thickness
varies over the bubble length for bubbles of finite length. Numerical
simulations \cite{giavedoni-numerical} and experimental studies
\cite{kreutzer-pressure-drop} showed a deviation from the $\Ca^{2/3}$ rule and Reynolds number 
effects on the film thickness for capillary numbers larger than $0.003$.
To consistently  predict a flow pattern for capillaries in
different ranges of parameters, simulations have been conducted and have been typically validated with the
small capillary numbers Bretherton problem.

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} also used the finite volume method to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. \citeauthor{wong-films} \cite{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes in the
slug cross sections and menisci appearance.
\citet{heil-bretherton} and \citet{ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using the finite element method. \citet{giavedoni-numerical} performed cross validation of the
finite element solution with previously published results.
The solutions were obtained for circular and planar cases.

@@

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for a relatively large capillary numbers, and
found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}.

Resolving the interface on a fine level for capillary numbers smaller than
$0.003$ is computationally expensive even in the two-dimensional case
(the amount of memory necessary to perform the simulation would be of the order
of tens or hundreds gigabytes).  This work is therefore based on a
comparison of the results obtained with the known data
validated for the Bretherton bubble flow.  In this paper, we present
techniques for initialization of the simulations and discuss optimal
parameter ranges for the binary liquid LB model.

@@

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), i=1\div8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \frac{\phi c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), i=1\div8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model. The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$ with
the sound speed parameter $c_s^2=1/3$.  The weights related to the
inclusion of the surface tension coefficient into the equations are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}=
-\partial_{\beta}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity,
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter, and $\tau=\frac{1}{\omega}$ and $\tau_{\phi}$
are the relaxation parameters of density and phase fields.

The system allows the separation of the liquid
phase with $\phi=1$ and a so-called gas phase with $\phi=-1$. The
relaxation time is taken as linearly dependent on the relaxation
times $\tau_{\mathrm{gas}}$ and $\tau_{\mathrm{liq}}$:
$\tau=\tau_{\mathrm{gas}}+\frac{\phi+1}{2}(\tau_{\mathrm{liq}}-\tau_{\mathrm{gas}})$. This allows
to change viscosity from the gas viscosity
$\nu_{\mathrm{gas}}=\frac{1}{3}\Bigl(\tau_{\mathrm{gas}}-\frac{1}{2}\Bigr)$ to the liquid viscosity
$\nu_{\mathrm{liq}}=\frac{1}{3}\Bigl(\tau_{\mathrm{liq}}-\frac{1}{2}\Bigr)$ while phase changes
accordingly.The surface tension in the framework of the binary liquid model is $\sqrt{\frac{8 k
A}{9}}$.

\section{Lattice Boltzmann benchmark}
To properly simulate and compare simulation results with the data
published in the literature, one needs to design a numerical benchmark addressing
the following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffuse nature of the interface in the multiphase model should not
influence the film thickness results.  The same applies to spurious
currents arising due to surface tension discretization, which
can influence mass transfer in the thin film, thereby compromising the
overall solution. Therefore, the diffuse interface should not noticeably affect the film
thickness.
 \item The wettability coefficient which in principle can control the
dynamic angle and the menisci appearance \cite{pagonabarraga-finger} should not
influence the overall solution.
 \item The free-energy model examined in this work is a single density, two
viscosities model. The classical Bretherton problem is a free-interface problem between liquid and
gas
where the film thickness is established by itself. Thus, the viscosity and the density ratios for
the
classical
Bretherton problem should
be large. However, in terms of the lattice Boltzmann binary liquid framework only the viscosity
ratio can be addressed. Our computations show that the results are consistent with other works and
inertial effects can be neglected.
  \item The Bretherton problem addresses a bubble flow driven by a pressure
difference, not a body force. However, to the best of our knowledge the pressure boundary
conditions are still not yet developed for the binary liquid model, because it involves the
treatment of associated gradients and Laplacians on the boundary.
This kind of pressure boundary conditions for the binary liquid model is currently
under development, and preliminary results show that the differences between flows driven by a body
force and by
pressure difference are small as far as the film
 thickness is concerned.  In this work we limit ourselves to the study
of body force driven flows, because of their simplicity and better numerical stability. This
implies that we can use the periodic boundary conditions in the streamwise direction. As soon as the
periodic boundary
conditions are applied, not a single bubble but a bubble train is simulated. In this case one
needs
to ensure that the distance between bubbles is large enough to exclude mutual bubble influence.
 \item Most simulations are done for a channel with circular cross section,
which is quite difficult to address in terms of lattice Boltzmann framework.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed specifically for curved boundaries,
the analysis of error introduced by the boundary conditions becomes an issue. Although the circular
axis-symmetric case can be simulated on a rectangular grid by introducing specific mass and force
terms in the continuity and Navier-Stokes equation \cite{halliday-circular}, the lattice Boltzmann
incorporation of force and mass terms is in the process of development and validation. Therefore,
the benchmark is chosen as a plane-symmetric two-dimensional case.  
\end{itemize}
Given all the concerns and challenges, the suggested lattice Boltzmann framework
is a two-dimensional flow driven by a body force.  An analytical
solution to this problem is not known for a wide range of capillary
numbers.
Though some authors \cite{sehgal-microchannel} use the gravity driven model to
validate the Bretherton film thickness, the latter should be used with caution
since the velocity of the bubble is different from the fluid velocity and the
bubble shape
is different in the front and rear menisci depending on Bond number ($\mathrm{Bo}=\frac{\Delta \rho g
l^2}{\gamma}$). \citet{wong-films}
analytically described the film thickness variation along a bubble varying from $\Ca^{1/2}$ to
$\Ca^{2/3}$. The bubble should be long enough to have the film thickness proportional to
$\Ca^{2/3}$. Therefore, we chose the bubble length to be $5$ channel heights. This number is shown
to be sufficient to give results consistent with the theory. The
channel length is taken to be $3$ bubble lengths (or $15$ channel heights)
to minimize influence of one bubble on another,
because of the periodicity of boundary conditions.
Figure \ref{fig:benchmark:sketch} is a sketch of the geometry
used for the benchmark.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/benchmark.eps}
\caption{The benchmark sketch. $\delta$ corresponds to film thickness. $\xi$ corresponds to the interface thickness. We take the microchannel length to be 15 times larger than its height.
\label{fig:benchmark:sketch}}
\end{figure}

\section{Results}
\subsection{The nondimensionalization and initialization procedure}
We now introduce the procedure of nondimensionalization, as well as
the initialization technique we used to determine the parameters necessary for
the simulation. The capillary number defined in Eq.~\ref{capillary:number:definition} governs the
interface thickness.
%The bubble velocity $U_{\mathrm{bubble}}$ is taken in the center of the bubble where the film
%thickness is examined.
While parameters such as $\mu_{\mathrm{liq}}$, $\gamma$ and the capillary number $\Ca$ are given
or defined through the model parameters, the bubble velocity $U_{\mathrm{bubble}}$ is not given
explicitly.
In our simulations we approximate $U_{bubble}$ with the
Poiseuille profile center velocity, which is given as:
\begin{equation}
\begin{aligned}
U_{\mathrm{bubble}} \approx U_{\mathrm{liq}}=\frac{H_{\mathrm{eff}}^2}{8
\mu_{\mathrm{liq}}}\frac{\mathrm{d}P}{\mathrm{d}x}\\
U_{\mathrm{bubble}}\approx
\frac{{(N_y-2)}^2}{\frac{8}{6}(\tau_{\mathrm{liq}}-\frac{1}{2})}\frac{\mathrm{d}P}{\mathrm{d}
x } ,
\end{aligned}
\end{equation}
where $N_y$ is the grid number in the vertical direction. As will be discussed further, the
effective channel width is defined as $H_{\mathrm{eff}}=N_y-2$ because of bounce-back nodes to
mimic no-slip conditions at the walls.
For the Poiseuille profile one can estimate the body force (pressure gradient) through the
capillary number and grid resolution:
\begin{equation}
\label{poiseuille:velocity:center}
\begin{aligned}
&\frac{\mathrm{d}P}{\mathrm{d}x}=\frac{8}{{(N_y-2)}^2}\gamma \Ca.
\end{aligned}
\end{equation}
In simulations the body force (\ref{poiseuille:velocity:center}) causes the bubble
velocity to differ from that implied by the Poiseuille relation because of the bubble presence in a
flow. That is
why an iteration procedure is needed to achieve the 
desired velocity and 
capillary number. 
However, the initialization body force given by (\ref{poiseuille:velocity:center}) is a good
starting approximation.
